---
title: "Quantitative Genomics and Genetics 2018"
subtitle: "Computer Lab 2"
author: "Zijun Zhao"
date: "2/8/2018"
output: html_document
---

-- R Markdown   
-- Functions   
-- For loops   
-- If/else statements   
-- Vector and Matrix calculation   

--------------------------------------------------------------------------------------

### 1. R Markdown

#### Standard text formating
- *italics*
- **bold**

#### Latex equations
- conditional probability: $Pr(X_1|X_2) = \frac{Pr(X_1 \cap X_2)}{Pr(X_2)}$
- matrix: $$A_{m,n} =
 \begin{pmatrix}
  a_{1,1} & a_{1,2} & \cdots & a_{1,n} \\
  a_{2,1} & a_{2,2} & \cdots & a_{2,n} \\
  \vdots  & \vdots  & \ddots & \vdots  \\
  a_{m,1} & a_{m,2} & \cdots & a_{m,n}
 \end{pmatrix}$$

#### Code chunks
```{r}
example.vector1 <- 1:10
mean(example.vector1)
```

Code options   

- `include = FALSE` prevents code and results from appearing in the finished file. R Markdown still runs the code in the chunk, and the results can be used by other chunks.   
- `echo = FALSE` prevents code, but not the results from appearing in the finished file. This is a useful way to embed figures.   

For more information on using R Markdown: http://rmarkdown.rstudio.com/lesson-1.html   

--------------------------------------------------------------------------------------

### 2. Functions

- We learned that a function is something that takes in an input and gives you an output.

- The input for an R function is written between "( )" of a function, and the output(s) are the result(s) returned by the function.

- R has many built in functions for commonly used methods in statistics.

```{r, comment=NA}
# Examples of built in functions 
example.vector1 <- c(5,2,3,7,1,1,2,9,9)

# a function that calculates the mean
mean(example.vector1) 

# a function to index specific values
which(example.vector1==3)

```

- We can also build custom functions. 

```{r, comment=NA}
# the syntax for declaring functions, note the {} after function()

log10_add <- function(input1,input2){   # all the inputs are specified within the ( )
  cat("This is a custom function \n")   
  cat("The inputs are = ",input1,input2,"\n")  # showing you the inputs 
  output <- log10(input1) + log10(input2)       # creating an output within the function
  cat("The output is = ",output,"\n")          # print the output
  return(output)                               # return specifies the output
}

# Now we can call our custom functions like this
log10_add(100,1000)

# Note that the variable output is not created in our workspace 
ls()

# in order to save the result of a function to a variable we have to assign it to a variable
test.output <- log10_add(100,1000)
test.output
```

**Question 1**

- Can you guess what is going to happen?

```{r, comment = NA, eval = FALSE}
x <- 11

test_function <- function(y){
  output <- x + y
  return(output)
}

test_function(2) 

```

```{r, comment = NA, eval = FALSE}
test_function2 <- function(x,y,z){
  output1 <- x + y + z 
  output2 <- x * y * z
  output3 <- x ^ y ^ z 
  
}

x <- test_function2(1,2,3) 

print(x)

```

```{r, comment = NA, eval = FALSE}
x <- 10 

test_function3 <- function(){
  x <- 20
}

test_function3() 

print(x)

```


#### Installing and Loading packages

- We can also use functions by installing published packages if somebody else did the hard work for us.

- We can install packages that are published on CRAN by using install.packages().

```{r}
# install.packages("ggplot2")
```

- Once the installation is complete, we to load the package into your current R session in order to use it.

```{r}
library(ggplot2)
require(ggplot2)
```

- Now we can use the functions from ggplot2

```{r}
x <- seq(1,10)
y <- sample(1:100,10,replace=T)
ggplot(data.frame(x,y), aes(x,y)) + geom_point()
```

--------------------------------------------------------------------------------------

### 3. for loops

- For loops are mainly used in cases where you want to do a task multiple times.

```{r}
N <- 3
for ( i in 1:N ){
    cat("Processing loop number = ",i,"\n")
}
```

- You can also create a loop within a loop

```{r}
for ( outer in 1:3 ){
    cat("Processing Outer Loop #", outer, "\n")
  for ( inner in 1:2 ){
      cat("Processing |_ Inner Loop #", inner, "\n")
    }
}
```

**Question2**

- What is the final value of N ?

```{r, eval = FALSE}
N <- 3
for( i in 1:N){
  cat("Processing loop = ", i, "\n")
  N <- N + 1
}
print(N)
```


--------------------------------------------------------------------------------------

### 4. If / else statements 

- By using if and else statements you can insert condition specific executions in your script

- The code inside an if statement will only be executed when the condition is TRUE

```{r, comment = NA, eval = FALSE}
if (condition) {
  do stuff
} else {
  do stuff
}

# OR you can add more levels by using else if

if(condition){
  do stuff
} else if (condition 2){
  do stuf
} else {
  do stuff
}
```

- Here is a simple example

```{r, comment = NA}
example.vector <- seq(1,25,by= 2)
# Loop over individual elements in example.vector
for( i in example.vector){
    if( i < 10 ){
        cat(i, "is smaller than 10 \n") 
    } else if ( 10 <= i & i <= 20){
        cat(i, "is in the interval [10,20] \n")
    } else {
        cat(i, "is larger than 20 \n")
    }
}
```


--------------------------------------------------------------------------------------

### 5. Vector and Matrix calculations

- If you want to modify each element of a vector by a scalar value you can use the math operations that we have learned last week.

```{r}
example.vector1
2 * example.vector1
1 + example.vector1
example.vector1 ^2
```

- If you are interested in the dot product of two vectors you have to use a special operator

```{r}
example.vector1 %*% example.vector1
```

- The same applies for matrices 

```{r}
example.matrix1 <- matrix(c(1,1,1,2,2,2), nrow = 2, ncol = 3, byrow= TRUE)
example.matrix1
2 * example.matrix1
example.matrix1 ^ 2
example.matrix1 - 1
```

- Here is how you can do matrix calculations

```{r, comment = NA}
# t() is transposing the matrix
example.matrix1 %*% t(example.matrix1)
# Note the dimensions 2 x 3 %*% 3 x 2  = 2 x 2 
```

- Here are some useful functions that can be used in matrix calculations 

```{r, comment = NA}
# creating a diagonal matrix with the first input as values on the diagonal
diag(2,nrow = 3)

# calculating the inverse of a matrix
A <- matrix(c(2,-3,1,0.5),nrow = 2)
solve(A)

# we can check this by 
A %*% solve(A) # which results in an identity matrix 

```

--------------------------------------------------------------------------------------

### Advanced Challenge

Write a function that takes a vector with Weill Cornell grades as input, and output a list with the following entries:

- "LetterGrade" should have the converted letter grades of each student in it.

- "Overview" should have the number of students for each Letter grades

- An entry for each letter grade with name of students that received the corresponding letter grade.

```{r}
scores <- sample(c("NP","P","LP","HP","H"), 20, replace = TRUE)
names(scores) <- paste0("student", 1:length(scores))
print(scores)
standard_letter_grades <- c("H" = "A", "HP" = "B", "LP" = "C", "P" = "D", "NP" = "F")
```

- Your code should also work when the conversion table changes (no hard coding of letter grades are allowed!)

- Your function should also be verbose, meaning that it should tell the user how many students are in the input in total, and the breakdown for each grade as well.

**Hints**

- Getting names of vectors 

```{r, comment =NA, eval = FALSE}
names(scores)
```

- Checking if value is equal to something

```{r, comment =NA, eval = FALSE}
value == 1
value == "A"
```

- Example output

```{r, comment = NA, echo = FALSE}
lab2_challenge <- function(input_scores, conversion_table){
  
  cat("Total number of students = ", length(input_scores), "\n")
  
  student_letter_grade <- unname(conversion_table[input_scores])
  names(student_letter_grade) <- names(input_scores)
  
  output_list <- list()
  output_list[["Overview"]] <- table(student_letter_grade)
  output_list[["LetterGrade"]] <- student_letter_grade
  for(grade in unique(conversion_table)){
    output_list[[grade]] <- names(student_letter_grade[student_letter_grade == grade]  )
    cat(length(output_list[[grade]]), "students received ", grade, "\n")
  }
  return(output_list)
}

lab2_challenge(scores, standard_letter_grades)

```















